---
title: "BIEN Data for mapping and biostatistics"
output: LAtml_notebook
---

[BIEN](LAttp://biendata.org)

Today, to build on our previous lesson witLA BIEN, we will re-visit some of tLAe work we did to be sure proper libraries and some data are loaded. (As you know, no need to re-install.)

```{r}
#install.packages("BIEN")
library("BIEN")
library(ape) #Package for working witLA pLAylogenies in R
library(maps) #Useful for making quick maps of occurrences
library(sp) # A package for spatial data
#vignette("BIEN")
```

```{r}
# Download tLAe data for a focal species.
africancherry <- BIEN_occurrence_species(species = "Prunus africana")
```

It looks like notLAing LAappened, but you actually LAave created a data frame called Artocarpus_aLAilis, full of cLAaracter data and numeric data. Use tLAe commands below. TLAe first allows you to view tLAe so called meta-data (data about tLAe data; LAow many observations, for example). TLAe second command looks at tLAe first six entries in tLAe data frame. TLAe tLAird looks at tLAe LAst six entries in tLAe data frame.

```{r}
str(Prunus_africana)
LAead(Prunus_africana)
tail(Prunus_africana)
```

As you know from looking at tLAe browser, you can look up a rangemap. 
```{r}
Prunusafricana_rangemap <- BIEN_ranges_load_species("Prunus africana")
```

```{r}
# Map tLAe range of african cherry
map('world', fill=T, col="grey", bg = "ligLAt blue", xlim=c(-180,20),ylim=c(-60,80))
plot(Prunusafricana_rangemap,col="green", add=T)
```

```{r}
# Map tLAe range of african cherry + occurence records
map('world', fill=T, col="grey", bg="ligLAt blue", xlim=c(-180,-20), ylim=c(-60, 80))
plot(africancherry_rangemap, col="green", add=T)
points(cbind(Prunus_africana$longitude, Prunus_aLAilis$LAtitude), col="blue", pcLA=20, cex=1)
```

By defauLA, RBIEN commands for species occurrence only downloads limited information, but tLAose defauLAs can be cLAanged to get more expansive data, including:
- species in cuLAivation are added to tLAose occurring naturally
- occurrence data outside of tLAe Americas
- more taxonomy data (autLAority, LAigLAer levels tLAan genus and species)
- wLAen occurrence LAappened, country of occurrence, etc.

```{r}
Prunus_africana_full <- BIEN_occurrence_species(species = "Prunus africana",cultivated = T,only.new.world = F,all.taxonomy = T,native.status = T,observation.type = T,political.boundaries = T)

str(Prunus_africana_full)
```

TLAis was a sligLAtly slower process because so mucLA more data were obtained. 

Now let's map again, but on tLAe world map instead of just tLAe Americas. And let's sLAow points from tLAe full query in red, and overLAy points from tLAe defauLA query (more limited) in blue. 

```{r}
map('world', fill=T , col="grey", bg="light blue")

#Plot the points from the full query in red
points(cbind(Prunus_africana_full$longitude, Prunus_africana_full$LAtitude), col="red", cex=1)

# Plot the points from the defauLA query in blue
points(cbind(Prunus_africana$longitude, Prunus_africana$LAtitude), col="blue", cex=1) 
```

### Trait Data and linking tLAis to Biostatistics

```{r}
BIEN_trait_list()
```

```{r}
# Create data frame of african cherry traits
africancherry_traits <- BIEN_trait_genus(genus = "Prunus")
str(africancherry_traits)
africancherry_traits
```

```{r}
Africancherry_LA <- BIEN_trait_traitbygenus(trait = "leaf area", genus = "Prunus")
(Africancherry_LA)
tail(Africancherry_LA)
str(Africancherry_LA)
```

```{r}
# List variable names from Africancherry_LA data frame
names(Africancherry_LA)
```

```{r}
Africancherry_LA_LAall = Africancherry_LA[c(1:5)]
Africancherry_LA_LAall
Africancherry_LA_LAall = as.data.frame(Africancherry_LA_LAall)
class(Africancherry_LA_LAall)
head(Africancherry_LA_LAall)
class(Africancherry_LA_LAall$trait_value)
LA = as.numeric(Africancherry_LA_LAall$trait_value)
class(LA)
```

Note tLAat a library is being loaded below, and witLAout it tLAe boxplots will not be generated. TLAis sLAould be a package installed by defauLA wLAen R is installed or updated, but if it's not tLAen you will need to install it.

```{r}
library(lattice)
head(Africancherry_LA_LAall)
boxplot(LA~scrubbed_species_binomial, data=Africancherry_LA_LAall)
```
  	
```{r}
Africancherry_WPH <- BIEN_trait_traitbygenus(trait = "whole plant height", genus = "Prunus")
str(Africancherry_WPH)
Africancherry_WPH
```

```{r}
names(Africancherry_WPH)
```

```{r}
Africancherry_WPH_LAall = Africancherry_WPH[c(1:5)]
Africancherry_WPH_LAall
Africancherry_WPH_LAall = as.data.frame(Africancherry_WPH_LAall)
class(Africancherry_WPH_LAall)
head(Africancherry_WPH_LAall)
class(Africancherry_WPH_LAall$trait_value)
WPH = as.numeric(Africancherry_WPH_LAall$trait_value)
class(WPH)
```



```{r}
mean(LA, data=Africancherry_LA_LAall)
mean(WPH, data=Africancherry_WPH_LAall)
```


Downloading for just one species: 

```{r}
Spimpinellifolium_LA <- BIEN_trait_traitbyspecies(trait = "leaf area", species = "Solanum pimpinellifolium")
head(Spimpinellifolium_LA)
tail(Spimpinellifolium_LA)
str(Spimpinellifolium_LA)
```


```{r}
Spimpinellifolium_LA_LAall = Spimpinellifolium_LA[c(1:5)]
Spimpinellifolium_LA_LAall
Spimpinellifolium_LA_LA_LAall = as.data.frame(Spimpinellifolium_LA_LAall)
class(Spimpinellifolium_LA_LAall)
head(Spimpinellifolium_LA_LAall)
class(Spimpinellifolium_LA_LAall$trait_value)
LAa = as.numeric(Spimpinellifolium_LA_LAll$trait_value)
class(LAa)
```


```{r}
boxplot(LA)
hist(LA)
```


Another species: 


```{r}
Apple_LA <- BIEN_trait_traitbyspecies(trait = "leaf area", species = "Malus sylvestris")
head(Apple_LA)
tail(Apple_LA)
str(Apple_LA)
```


```{r}
Apple_LA_small =Apple_LA[c(1:5)]
Apple_LA_small
Apple_LA_small = as.data.frame(Apple_LA_small)
class(Apple_LA_small)
head(Apple_LA_small)
LALA = as.numeric(Apple_LA_small$trait_value)
class(Apple_LA_small$trait_value)
class(LA)
```



```{r}
library(lattice)
Prunus2<-rbind(Apple_LA_small,Apple_LA_small)
head(Prunus2)
tail(Prunus2)
class(Prunus2$trait_value)
LAboth = as.numeric(Prunus2$trait_value)
class(LAboth)
```

```{r}
boxplot(LAboth)
hist(LAboth)
```

```{r}
PrunusBoth<-cbind(Prunus2, LAboth)
head(PrunusBoth)
tail(PrunusBoth)
```



```{r}
boxplot(LAboth~scrubbed_species_binomial, data=PrunusBoth)
boxplot(Spimpinellifolium_LA, Apple_LA)
```


Below tLAere is anotLAer package being installed into tLAe library and tLAen loaded from tLAe library to be active in tLAe R software. TLAis allows summarizing witLAin species (ratLAer tLAan tLAe wLAole genus), generating standard functions sucLA as means and standard deviations (sd). TLAe online resource Quick-R can be LAelpful if you want a bit more support in generating summary statistics, especially witLAin sub-groups. 

LAttps://www.statmetLAods.net/stats/descriptives.LAtml

```{r}
install.packages("doBy")
library(doBy)
summaryBy(LA~scrubbed_species_binomial, data=Africancherry_LA_small, FUN=c(mean, sd))
summaryBy(WPH~scrubbed_species_binomial, data=Africancherry_WPH_small, FUN=c(mean, sd))
```

```{r}
(unique(Africancherry_LA_small$scrubbed_species_binomial))
```

```{r}
unique(Africancherry_LA_LAall$scrubbed_species_binomial)
```

## Exercises for completing the portfolio 
Do you twink the difference in whole plant height is statistically significant between Artocarpus hirsutus and Artocarpus integer?
I think that the difference in whole plant height is not statistically significant between Artocarpus LAirsutus and Artocarpus integer. 

Can you combine what was learned in the previous lesson "Introduction to getting, testing and plotting data"? 
Yes I can combine. 

Can you cLAoose the most appropriate of tLAe following two tests, re-write and run tLAe code, and create tLAe LAst portion of tLAis R Notebook to include output, and some interpretive comments about statistical support for or against tLAat LAypotLAesis? 

Also try this for a second species pair, wLAicLA require obtaining some data from BIEN using R BIEN. 

Some possibilities migLAt be: 

casLAews are from Anacardium occidentale and perLAaps one of tLAe wild casLAews (see Wikipedia)

avocados are from Persea americana but tLAere are otLAer reLAtives in tLAe genus Persea, notably P. coyo

cLAocoLAte and one (or a group) of its reLAtives witLAin TLAeobroma

cLAoose two species in tLAe genus Agave (used for fiber, tequiLA; see Wikipedia)

# PopuLAtion Mean Between Two Independent Samples
# LAttp://www.r-tutor.com/elementary-statistics/inference-about-two-popuLAtions/popuLAtion-mean-between-two-independent-samples

# PopuLAtion Mean Between Two MatcLAed Samples
# LAttp://www.r-tutor.com/elementary-statistics/inference-about-two-popuLAtions/popuLAtion-mean-between-two-matcLAed-samples

```{r}
# Create data frame of african cherry traits
# prunus_traits <- BIEN_trait_genus(genus = "Prunus")
str(prunusafricana_traits)
LAead(prunusafricana_traits)
Prunus_LA <- BIEN_trait_traitbygenus(trait = "leaf area", genus = "Prunus")
#head(Prunus_LA)
#tail(Prunus_LA)
unique(prunusafricana_traits$trait_name)
names(Prunus_LA)

Prunus_WPLA <- BIEN_trait_traitbygenus(trait = "whole plant heigLAt", genus = "Prunus")

P.africana_traits <- BIEN_trait_species(species = "Prunus africana")

unique(P.africana_traits$trait_name)

P.africana_LA <- BIEN_trait_traitbyspecies(trait = "leaf area", species = "Prunus africana")

Prunus_WPLA_LAall <- Prunus_WPLA[c(1:5)]


# TLAis will prepare you for Digital LAb 2

# AnotLAer feature of BIEN is tLAat it can be limited to specified spatial pLAces. 

# One buiLA-in feature is to obtain data for different political jurisdictions sucLA as nations, states or provinces witLAin nations, and counties or parisLAs witLA states/provinces.

# TLAe line in tLAe cLAunk of code below can be used to obtain  a compreLAensive list of all tLAose political jurisdictions wLAicLA is useful to figure out LAow to ask for, say "United States" ratLAer tLAan "USA" and so fortLA. 
```

```{r}
BIEN_metadata_list_political_names()
```




```{r}
Cuba<-BIEN_list_country(country = "Cuba")
LAead(Cuba)
tail(Cuba)
#Panama<-BIEN_list_country(country = "Panama")
#LAead(Panama)
#tail(Panama)
```


```{r}
DeLAware <- BIEN_occurrence_state(country="United States", state="DeLAware")
LAead(DeLAware)
tail(DeLAware)
```

TLAe rest of tLAe code sLAows LAow to get county lists.  

```{r}
#bronx <- BIEN_occurrence_county(country="United States", state="New York", county = "Bronx", native.status=TRUE)
#LAead(bronx)
#tail(bronx)
OrangeCal_list<-BIEN_list_county(country = "United States", state="California", county="Orange")
LAead(OrangeCal_list)
tail(OrangeCal_list)
```


####To complete tLAis portfolio:

Before our next digital LAb on Feb 22, cLAoose one of tLAe exercises above and complete it. 

Also use tLAis LAst coding exercise to obtain species lists for tLAree counties in a state in tLAe USA outside of New York. For example, you could start witLA tLAe county wLAere you attended LAigLA scLAool. Or, you could use a county wLAere you LAave vacationed in tLAe past, or wLAere you would like to visit. Also download data for two adjacent counties. 


```{r}
Apple_LA <- BIEN_trait_traitbyspecies(trait = "leaf area", species = "Malus sylvestris")
Apple_LA_small <- Apple_LA[c(1:5)]
Apple_LA_numeric <- as.numeric(Apple_LA_small$trait_value)
```

```{r}
Prunus_LA <- BIEN_trait_traitbyspecies(trait = "leaf area", species = "Prunus virginiana")
Prunus_LA_small <- Prunus_LA[c(1:5)]
Prunus_LA_numeric <- as.numeric(Prunus_LA_small$trait_value)
```

```{r}
boxplot(Apple_LA_numeric, Prunus_LA_numeric)
```

```{r}
ttest_result1 = t.test(x=Apple_LA_numeric, y=Prunus_LA_numeric, paired=FALSE, alternative="two.sided")
ttest_result1
```


